Measuring the Type Ia supernova delay time distribution via host galaxy rates in the Dark Energy Survey

### 2021-01-14 draft.tex

Given the vast range of timescales to which the DTD is sensitive, it is non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SNe Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations heavily imply that SNe Ia rates are dependent on recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Smith2012}. The DTD is thus approximated by a two-component ("$A + B$") model, where $A$ and $B$ are the normalizations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008}. This technique is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Graur2014,Maoz2014}. Conversely, it is also possible to estimate the SFH of an individual galaxy through the modelling of its stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012}, \citet{Graur2013} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy. and has led to results that suggest a DTD power law with $\beta$ consistent with $-1$ \citep[e.g.][]{Maoz2011,Graur2013}. \citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the SFHs of galaxies.

@@

%\begin{figure}
%    \centering
%    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_split_sfr.png}
%    \caption{The rate per galaxy of SNe Ia as a function of stellar mass for moderately and highly star-forming galaxies, as well as passive galaxies.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
%    \label{fig:rate_raw_split_sfr}
%\end{figure}
%%% MODEL %%%
\section{Modelling the rate of type Ia supernovae}
 \label{sec:method}
We represent the SN Ia DTD by a power law. We consider the DTD effective after some "prompt time" $t_{\mathrm{p}}$, before which its value is set to 0. This is generally interpreted as the time taken for WDs to form after the burst of star formation. The DTD is thus:
\begin{equation}
 \Phi(\tau) = \left\{
    \begin{array}{@{}cc@{}}
    0 , & \tau < t_{\mathrm{p}} \\
    A\left(\frac{\tau}{\mathrm{Gyr}}\right)^{\beta} , & \tau \geq t_{\mathrm{p}}
    \end{array}\right.
        \label{eq:dtd}
\end{equation} 
where $\tau$ is the time since a burst of star formation, and $A$ is a normalisation at $\tau=1$ Gyr. If all stars in a galaxy were formed at a single epoch $t_f$, the DTD would describe the rate of SNe at an observation time $\tau = t_0$. However in reality galaxies are formed by the gradual build up of stellar mass over several epochs of star formation -- the distribution of this mass build up is known as the star-formation history (SFH). The rate of SNe Ia is thus the sum of the DTD evaluated for each epoch of star formation, multiplied by the stellar mass formed in that epoch. Mathematically this is represented by the convolution of the DTD and SFH:

@@

\section{Results}
\label{sec:rates}
The rate of SNe per galaxy ($R_{\mathrm{G}}$) per year in a transient survey can be approximated by the equation:
\begin{equation}
    R_{\mathrm{G}} = \frac{N_{\mathrm{SN}}}{N_{\mathrm{G}}} \frac{V_{\mathrm{G}}}{V_{\mathrm{SN}}}\frac{1}{T_{\mathrm{SN  survey}}}\,,
\label{eq:rate1}
\end{equation}
where $N_{\mathrm{SN}}$ and $N_{\mathrm{G}}$ are the respective number of SNe and galaxies detected, $V_{\mathrm{SN}}$ and $V_{\mathrm{G}}$ the volumes from which the SNe and G samples were taken, and $T$ the duration of the SN survey in years.

\subsection{Dependence of the SN Ia rate on stellar mass \label{subsec:rates_mass}}

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we employ Hamiltonian Monte-Carlo (HMC), which we implement via the \texttt{pystan} package and describe in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.70\pm0.02$, which is broadly consistent with the values found in SNLS and SDSS by \citet{Sullivan2006} and \citet{Smith2012} respectively. It is evident that the linear fit is not consistent with the data across the full range of stellar mass considered. In particular, there appear to be two locations of interest: firstly, the slope is steeper at masses lower than $10^9$ M$_{\odot}$; secondly there is an apparent flattening and possibly a turnover at masses around and above $10^11$ M$_{\odot}$. The most likely cause of these divergences is because the SN rate is driven by the SFH which is not a linear function of stellar mass (see Section \ref{sec:method}). Second order effects may also be important, and we address these in Section \ref{sec:discussion}.

@@

\subsection{Constraints on the SN Ia delay time distribution}
\label{subsec:results_dtd}

In Section \ref{sec:method} we showed how the rate of SNe Ia is driven by the convolution of the SFH of galaxies and the SN Ia DTD (Eq. \ref{eq:galaxy_rate}). We then prescribed a model to infer the mean SFHs of galaxies of any given stellar mass. In this section, we measure the DTD by forward modelling it through Eq. \ref{eq:galaxy_rate} at the stellar masses corresponding to the centres of the bins in Fig. \ref{fig:rate_raw}.

@@

SNe Ia do not occur instantaneously after an episode of star formation. Instead stars with zero-age main-sequence (ZAMS) masses less than $8 \mathrm{M}_{\odot}$ take time to evolve along the main sequence and form WDs and this time is not well constrained. In many works this ``prompt time" $t_{\mathrm{p}}$ is fixed to some value expected to be the minimum time for a star to evolve off the MS and become a WD, such as 40 Myr \citep{Maoz2012,Graur2014,Graur2018}. Other studies have left $t_{\mathrm{p}}$ as a free parameter \citep{Heringer2019,Asier2020}. We perform two fits: one with $t_{\mathrm{p}}$ fixed at 40 Myr, and one with it as a free parameter. We fit $t_{\mathrm{p}}$ in log space and adopt a Gaussian prior with hyper-parameters $p(\log(t_{\mathrm{p}})) \sim \mathcal{N}(-1.3\,,0.2)$, which corresponds to being centred around 50 Myr. Choosing this regime to fit allows more prior weight to be placed on shorter prompt times as per the majority of the literature.

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. The inclusion of $t_{\mathrm{p}}$ in the fit produces results for all three parameters that are significantly different than in the fixed case, suggesting either that the prompt time is different to 40 Myr, or that the method is incapable of constraining these three parameters simultaneously.

\section{Discussion}
\label{sec:discussion}

>>>>

Given the vast range of timescales to which the DTD is sensitive, it is non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SNe Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations heavily imply that SNe Ia rates are dependent on recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Smith2012}. The DTD is thus approximated by a two-component ("$A + B$") model, where $A$ and $B$ are the normalizations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2018, Frohmaier2019}. This technique is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Graur2014,Maoz2014}. Conversely, it is also possible to estimate the SFH of an individual galaxy through the modelling of its stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012}, \citet{Graur2013} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy. and has led to results that suggest a DTD power law with $\beta$ consistent with $-1$ \citep[e.g.][]{Maoz2011,Graur2013}. \citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the SFHs of galaxies.

@@

%\begin{figure}
%    \centering
%    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_split_sfr.png}
%    \caption{The rate per galaxy of SNe Ia as a function of stellar mass for moderately and highly star-forming galaxies, as well as passive galaxies.}%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
%    \label{fig:rate_raw_split_sfr}
%\end{figure}
%%% MODEL %%%

\section{The per-galaxy rate of type Ia supernovae}
\label{sec:rates}
The rate of SNe per galaxy ($R_{\mathrm{G}}$) per year in a transient survey can be approximated by the equation:
\begin{equation}
    R_{\mathrm{G}} = \frac{N_{\mathrm{SN}}}{N_{\mathrm{G}}} \frac{V_{\mathrm{G}}}{V_{\mathrm{SN}}}\frac{1}{T_{\mathrm{SN  survey}}}\,,
\label{eq:rate1}
\end{equation}
where $N_{\mathrm{SN}}$ and $N_{\mathrm{G}}$ are the respective number of SNe and galaxies detected, $V_{\mathrm{SN}}$ and $V_{\mathrm{G}}$ the volumes from which the SNe and G samples were taken, and $T$ the duration of the SN survey in years.

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/beta_A_corner.png}
    \caption{Posterior distributions for DTD slope $\beta$ and normalisation $A$. %The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:corner_beta_norm}}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/beta_A_tp_corner.png}
    \caption{Posterior distributions for DTD parameters $\beta$, $A$, and $t_{\mathrm{p}}$. %The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:corner_beta_norm_tp}}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm.png}
    \caption{Rate per galaxy of SNe Ia as a function of stellar mass (green points) along with the prediction from the best fitting DTD parameters $\beta$, $A$, and $t_{\mathrm{p}}$ forward modelled through Eq. \ref{eq:galaxy_rate}. Samples from the posterior distribution of the model log rate are drawn in cyan while the black dashed line is the mean of the posterior.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted}}
\end{figure}

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we employ Hamiltonian Monte-Carlo (HMC), which we implement via the \texttt{pystan} package and describe in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.70\pm0.02$, which is broadly consistent with the values found in SNLS and SDSS by \citet{Sullivan2006} and \citet{Smith2012} respectively. It is evident that the linear fit is not consistent with the data across the full range of stellar mass considered. In particular, there appear to be two locations of interest: firstly, the slope is steeper at masses lower than $10^9$ M$_{\odot}$; secondly there is an apparent flattening and possibly a turnover at masses around and above $10^{11}$ M$_{\odot}$. The most likely cause of these divergences is because the SN rate is driven by the SFH which is not a linear function of stellar mass (see Section \ref{sec:method}). Second order effects may also be important, and we address these in the following sections.

\section{Modelling the per-galaxy rate of type Ia supernovae}
 \label{sec:model}

In Section \ref{sec:rates} and Fig. \ref{fig:rate_raw} we showed that the SN Ia rate per galaxy as a function of stellar mass is approximated by, but not perfectly described by, a power-law (and thus a linear relationship in log space). In this section, we introduce a physical model in order to better fit the observed rate vs stellar mass relationship.

We begin by considering the delay time distribution of SNe Ia.
We represent the SN Ia DTD by a power law, which we consider effective after some "prompt time" $t_{\mathrm{p}}$, before which its value is set to 0. This is generally interpreted as the time taken for WDs to form after the burst of star formation. The DTD is thus:
\begin{equation}
 \Phi(\tau) = \left\{
    \begin{array}{@{}cc@{}}
    0 , & \tau < t_{\mathrm{p}} \\
    A\left(\frac{\tau}{\mathrm{Gyr}}\right)^{\beta} , & \tau \geq t_{\mathrm{p}}
    \end{array}\right.
        \label{eq:dtd}
\end{equation} 
where $\tau$ is the time since a burst of star formation, and $A$ is a normalisation at $\tau=1$ Gyr. If all stars in a galaxy were formed at a single epoch $t_f$, the DTD would describe the rate of SNe at an observation time $\tau = t_0$. However in reality galaxies are formed by the gradual build up of stellar mass over several epochs of star formation -- the distribution of this mass build up is known as the star-formation history (SFH). The rate of SNe Ia is thus the sum of the DTD evaluated for each epoch of star formation, multiplied by the stellar mass formed in that epoch. Mathematically this is represented by the convolution of the DTD and SFH:

@@

@@

\begin{table}
	\centering
	\caption{This is an example table. Captions appear above each table.
	Remember to define the quantities, symbols and units used.}
	\label{tab:dtd_results}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		 &$\beta$ & $A$ & $t_{\mathrm{p}}$\\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ & Gyr \\
		\hline
		Fixed $t_{\mathrm{p}}$ & $-1.04$ &  $1.7$ & -\\
		Free $t_{\mathrm{p}}$ & $-0.99 \pm 0.09$ & $2.1 \pm 0.2$ & $0.09 \pm 0.05$\\
		\hline
	\end{tabular}
\end{table}

\subsection{Constraining the SN Ia delay time distribution}
\label{subsec:results_dtd}

At the beginning of this Section, we showed how the rate of SNe Ia is driven by the convolution of the SFH of galaxies and the SN Ia DTD (Eq. \ref{eq:galaxy_rate}). We then prescribed a model to infer the mean SFHs of galaxies of any given stellar mass. Here, we measure the DTD by forward modelling it through Eq. \ref{eq:galaxy_rate} at the stellar masses corresponding to the centres of the bins in Fig. \ref{fig:rate_raw}.

@@

SNe Ia do not occur instantaneously after an episode of star formation. Instead stars with zero-age main-sequence (ZAMS) masses less than $8 \mathrm{M}_{\odot}$ take time to evolve along the main sequence and form WDs and this time is not well constrained. In many works this ``prompt time" $t_{\mathrm{p}}$ is fixed to some value expected to be the minimum time for a star to evolve off the MS and become a WD, such as 40~Myr \citep{Maoz2012,Graur2014,Graur2018}. Other studies have left $t_{\mathrm{p}}$ as a free parameter \citep{Heringer2019,Asier2020}. We perform two fits: one with $t_{\mathrm{p}}$ fixed at 40 Myr, and one with it as a free parameter. We fit $t_{\mathrm{p}}$ in log space and adopt a Gaussian prior with hyper-parameters $p(\log(t_{\mathrm{p}})) \sim \mathcal{N}(-1.3\,,0.2)$, which corresponds to being centred around 50~Myr. Choosing this regime to fit allows more prior weight to be placed on shorter prompt times as per the majority of the literature.

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. In Fig. \ref{fig:rate_fitted} we show the SN rate per galaxy predicted by the model assuming the best fitting DTD parameters from Table \ref{tab:dtd_results} compared to the data as presented in Section \ref{sec:rates}. The model provides a good fit across a wide range of stellar mass. The model clearly diverges from the simple log-linear fit, describing better the suppressed rate at the low and high ends of the stellar mass range, as well as the enhanced rate around $10^{10}$ M$_{\odot}$. This improvement provides evidence to support the DTD*SFH model. However, the model prediction still diverges from the data at low ($<10^9$ M$_{\odot}$) and high ($>10^{11.25}$ M$_{\odot}$), suggesting that these data cannot be explained by the our model, but are driven by further processes that we have not included. We discuss these in Section \ref{subsec:discussion}.

\subsection{Comparison to previous DTD measurements \label{subsec:compare_dtd}}

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
The index of the DTD power law measured here is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2018,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}.

\subsubsection{The DTD normalisation \label{subsubsec:compare_A}}
The normalisation of the DTD is less frequently reported. \citet{Graur2013} found a value of $7 \times 10^{-14}$~M$_{\odot}^{-1}$~yr$^{-1}$ using the SFH technique, whereas \citet{Heringer2019} report $7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. These results lie either side of our measured value. \citet{Heringer2019} note that the normalisation recovered from the SFHR method is sensitive to the assumed prompt time $t_{\mathrm{p}}$, which we also find using our method.

\subsubsection{The SN Ia prompt time \label{subsubsec:compare_tp}}

The majority of works in the field have constrained the DTD power-law slope, and many also estimate its normalisation. Conversely, few have attempted to constrain the time after a burst of star-formation at which SNe Ia begin to explode -- the prompt time $t_{\mathrm{p}}$. In most cases, $t_{\mathrm{p}}$ has been fixed at some fiducial value such as 40~Myr, which is derived from the lifetime of 8~M$_{\odot}$ stars.

\subsection{Second order processes affecting the supernova rate \label{subsec:discussion}}

\subsubsection{Stochastic SFHs}
